A tutorial on Bayesian and Frequentist Event History Analyses for psychological time-to-event data

### 2024-09-22 manuscript/manuscript.Rmd
update manuscript

In experimental psychology, it is still standard practice to analyse response times (RTs), saccade latencies, and fixation durations by calculating average performance across a series of trials. However, differences in mean RT conceal when an experimental effect starts, how long it lasts, how it evolves over increasing waiting time, and whether its onset is time-locked to other events. Such information is useful not only for interpretation of the effects, but also for cognitive psychophysiology and computational model selection [@panisAnalyzingResponseTimes2020].
As a simple illustration, Figure 1 shows three examples of how an observed difference in mean response times (RTs) between two experimental conditions conceals differences in the shapes of the underlying RT distributions.

@@

In each example, the mean RT is lower in condition 2 compared to condition 1. However, the distributions in the first example show that the effect starts around 200 ms and is gone by 600 ms. In the second example, the effect starts around 400 ms and is gone by 800 ms. And in the third example, the effect reverses around 550 ms.

@@

We then provide four different tutorials, each of which is written in R code and publicly available on Github and the Open Science Framework (OSF). The tutorials provide hands-on, concrete examples of key parts of the analytical process, so that others can apply the analyses to their own time-to-event data sets. In Tutorial 1 we illustrate how to calculate the descriptive statistics for a published data set when there is one independent variable. The descriptive statistics are plotted, and we comment on their interpretation. In Tutorial 2 we illustrate how one can fit Bayesian hazard models to the data. After selecting the best of four models, we plot the model-based effects and the model fits for a few subjects. In Tutorial 3 we illustrate how to fit hazard models in a frequentist framework. We compare the model-based effects between Bayesian and frequentist approaches. In Tutorial 4 we illustrate how to calculate the descriptive statistics when there are two independent variables.

@@

Plot the effects of congruent and incongruent for each time bin for the selected model.

Figure 5

Plot the model-based hazard and survivor functions.

Figure 6

@@

In this third tutorial we illustrate how to fit a multilevel hazard regression model in the frequentist framework, for the data set used in the first tutorial.

model and comparison and note on fitting

Figure 7

@@

One important limitation is that you need a relatively large number of trials per condition to estimate the hazard function with high temporal resolution. There is a tradeoff between the number of trials per condition and the temporal resolution (i.e., bin width) of the hazard function. We therefore recommend to design as many trials as possible per experimental condition given the available resources.
Another limitation of multilevel hazard modeling is that it might take a long time to estimate the parameters.

@@

The hazard models in this tutorial assume that there is one event of interest. However, in certain situations, more than one event of interest might exist. For example, ... - multi-state hazard models refs

>>>>

In experimental psychology, it is still standard practice to analyse response times (RTs), saccade latencies, and fixation durations by calculating average performance across a series of trials. However, differences in mean RT conceal when an experimental effect starts, how long it lasts, how it evolves over increasing waiting time, and whether its onset is time-locked to other events. Such information is useful not only for interpretation of the effects, but also for cognitive psychophysiology and computational model selection [@panisAnalyzingResponseTimes2020].

As a simple illustration, Figure 1 shows three examples of how an observed difference in mean response times (RTs) between two experimental conditions conceals differences in the shapes of the underlying RT distributions. In each example, the mean RT is lower in condition 2 compared to condition 1. However, the distributions in the first example show that the effect starts around 200 ms and is gone by 600 ms. In the second example, the effect starts around 400 ms and is gone by 800 ms. And in the third example, the effect reverses around 550 ms.

@@

@@

We then provide four different tutorials, each of which is written in R code and publicly available on Github and the Open Science Framework (OSF). The tutorials provide hands-on, concrete examples of key parts of the analytical process, so that others can apply the analyses to their own time-to-event data sets. In Tutorial 1 we illustrate how to calculate the descriptive statistics for a published data set when there is one independent variable. The descriptive statistics are plotted, and we comment on their interpretation. In Tutorial 2 we illustrate how one can fit Bayesian hazard models to the data. We discuss possible link functions, and plot the model-based effects of our predictors of interest. In Tutorial 3 we illustrate how to fit hazard models in a frequentist framework. Even though both frameworks generate similar parameter estimates, we note that model convergence is often not obtained in the frequentist framework. In Tutorial 4 we illustrate how to calculate the descriptive statistics when there are two independent variables.

@@

Figure 5 shows the effects of congruent and incongruent primes relative to neutral primes, for each time bin in trial number 1000 for the selected model.

(ref:descr-fig5-caption) 50/80/95 percentile intervals of the draws from the posterior distributions representing the effect of congruent and incongruent primes relative to neutral primes in trial number 1000.

```{r plot-prime-effects, fig.cap = "(ref:descr-fig5-caption)", out.width='80%'}
knitr::include_graphics("../Tutorial_2_Bayesian/figures/M4effects_con_incon_trial1000.png")
```

Figure 6 shows the model-based hazard functions for each prime type for participant 6, in trial 500, 1000, and 1500.

(ref:descr-fig6-caption) Model-based hazard functions for participant 6 in trial 500, 1000, and 1500.

```{r plot-hazard-subject6, fig.cap = "(ref:descr-fig6-caption)", out.width='80%'}
knitr::include_graphics("../Tutorial_2_Bayesian/figures/M4effects_subject6.png")
```

@@

In this third tutorial we illustrate how to fit a multilevel hazard regression model in the frequentist framework, for the data set used in the first tutorial. For illustration purposes, we only fitted model M3 using the function glmer() from the package lme4.

\scriptsize 
```{r fit-model, eval=F, echo=T}
model_M3_f <- glmer(event ~ 1 + condition*period_9 + 
                           condition*I(period_9^2) + 
                           condition*I(period_9^3) +
                           trial_c*period_9 + 
                           trial_c*I(period_9^2) + 
                           trial_c*I(period_9^3) +
                           (1 + condition*period_9 +
                           condition*I(period_9^2) +
                           condition*I(period_9^3) +
                           trial_c*period_9 + 
                           trial_c*I(period_9^2) + 
                           trial_c*I(period_9^3) | pid),

              # control parameters, data set, and complementary log-log link function
              control = glmerControl(optimizer = c("nlminbwrap"),
                                     optCtrl = list(maxfun=10000000)), 
              data=M3_data, 
              family=binomial(link="cloglog"))
```
\normalsize

In Figure 7 we compere the parameter estimates of model M3 from brms() with those of glmer().

(ref:descr-fig7-caption) Parameter estimates for model M3 from brms() and glmer().

```{r plot-comparison, fig.cap = "(ref:descr-fig7-caption)", out.width='80%'}
knitr::include_graphics("../Tutorial_3_Frequentist/comparison.png")
```

Figure 7 confirms that the parameter estimates from both Bayesian and frequentist models are pretty similar. However, the random effects structure of model M3 was already too complex for the frequentist model as it did not converge and resulted in a singular fit. This is of course one of the reasons why Bayesian modeling has become so popular in the last years.

@@

Compared to the orthodox method -- comparing means with ANOVA --, the most important limitation of multilevel hazard modeling is that it might take a long time to estimate the parameters.
Another issue that might initially look as a limitation is that you need a relatively large number of trials per condition to estimate the hazard function with high temporal resolution. However, as nature does not reveal itself easily, obtaining insight into behavioral dynamics simply requires more data per condition than is usually collected under the orthodox method. In general, there is a tradeoff between the number of trials per condition and the temporal resolution (i.e., bin width) of the hazard function; We therefore recommend to design as many trials as possible per experimental condition given the available resources.

@@

The hazard models in this tutorial assume that there is one event of interest. For RT data, this event consitutes a single transition between an "idle" state and a "responded" state. However, in certain situations, more than one event of interest might exist. For example, an individual might transition back and forth between a "healthy" state and a "depression" state, before being absorbed in a final "death" state. When you have data on the timing of these transitions, one can apply multi-state models which generalize survival analysis to transitions between three or more states (refs).
Also, the predictor variables in this tutorial are time-invariant, i.e., their value did not change over the course of a trial. Another extension is to include time-varying predictors, i.e., predictors whose value can change across the time bins within a trial.

### 2024-09-22 manuscript/manuscript.tex
update manuscript

In experimental psychology, it is still standard practice to analyse response times (RTs), saccade latencies, and fixation durations by calculating average performance across a series of trials. However, differences in mean RT conceal when an experimental effect starts, how long it lasts, how it evolves over increasing waiting time, and whether its onset is time-locked to other events. Such information is useful not only for interpretation of the effects, but also for cognitive psychophysiology and computational model selection (Panis, Schmidt, Wolkersdorfer, \& Schmidt, 2020).
As a simple illustration, Figure 1 shows three examples of how an observed difference in mean response times (RTs) between two experimental conditions conceals differences in the shapes of the underlying RT distributions.

@@

We used R (Version 4.4.0; R Core Team, 2024)\footnote{We, furthermore, used the R-packages \emph{bayesplot} (Version 1.11.1; Gabry, Simpson, Vehtari, Betancourt, \& Gelman, 2019), \emph{brms} (Version 2.21.0; Bürkner, 2017, 2018, 2021), \emph{citr} (Version 0.3.2; Aust, 2019), \emph{cmdstanr} (Version 0.8.1.9000; Gabry, Češnovar, Johnson, \& Bronder, 2024), \emph{dplyr} (Version 1.1.4; Wickham, François, Henry, Müller, \& Vaughan, 2023), \emph{forcats} (Version 1.0.0; Wickham, 2023a), \emph{ggplot2} (Version 3.5.1; Wickham, 2016), \emph{lme4} (Version 1.1.35.5; Bates, Mächler, Bolker, \& Walker, 2015), \emph{lubridate} (Version 1.9.3; Grolemund \& Wickham, 2011), \emph{Matrix} (Version 1.7.0; Bates, Maechler, \& Jagan, 2024), \emph{nlme} (Version 3.1.166; Pinheiro \& Bates, 2000), \emph{papaja} (Version 0.1.2.9000; Aust \& Barth, 2023), \emph{patchwork} (Version 1.2.0; Pedersen, 2024), \emph{purrr} (Version 1.0.2; Wickham \& Henry, 2023), \emph{RColorBrewer} (Version 1.1.3; Neuwirth, 2022), \emph{Rcpp} (Eddelbuettel \& Balamuta, 2018; Version 1.0.12; Eddelbuettel \& François, 2011), \emph{readr} (Version 2.1.5; Wickham, Hester, \& Bryan, 2024), \emph{RJ-2021-048} (Bengtsson, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021), \emph{standist} (Version 0.0.0.9000; Girard, 2024), \emph{stringr} (Version 1.5.1; Wickham, 2023b), \emph{tibble} (Version 3.2.1; Müller \& Wickham, 2023), \emph{tidybayes} (Version 3.0.6; Kay, 2023), \emph{tidyr} (Version 1.3.1; Wickham, Vaughan, \& Girlich, 2024), \emph{tidyverse} (Version 2.0.0; Wickham et al., 2019), and \emph{tinylabels} (Version 0.2.4; Barth, 2023).} for all reported analyses. Web links are printed in bold. The content of the tutorials is mainly based on Allison (2010), Singer and Willett (2003), McElreath (2018), Kurz (2023a), and Kurz (2023b).

@@

Plot the effects of congruent and incongruent for each time bin for the selected model.

Figure 5

Plot the model-based hazard and survivor functions.

Figure 6

@@

In this third tutorial we illustrate how to fit a multilevel hazard regression model in the frequentist framework, for the data set used in the first tutorial.

model and comparison and note on fitting

Figure 7

@@

The hazard models in this tutorial assume that there is one event of interest. However, in certain situations, more than one event of interest might exist. For example, \ldots{} - multi-state hazard models refs

>>>>

In experimental psychology, it is still standard practice to analyse response times (RTs), saccade latencies, and fixation durations by calculating average performance across a series of trials. However, differences in mean RT conceal when an experimental effect starts, how long it lasts, how it evolves over increasing waiting time, and whether its onset is time-locked to other events. Such information is useful not only for interpretation of the effects, but also for cognitive psychophysiology and computational model selection (Panis, Schmidt, Wolkersdorfer, \& Schmidt, 2020).

As a simple illustration, Figure 1 shows three examples of how an observed difference in mean response times (RTs) between two experimental conditions conceals differences in the shapes of the underlying RT distributions. In each example, the mean RT is lower in condition 2 compared to condition 1. However, the distributions in the first example show that the effect starts around 200 ms and is gone by 600 ms. In the second example, the effect starts around 400 ms and is gone by 800 ms. And in the third example, the effect reverses around 550 ms.

@@

We used R (Version 4.4.0; R Core Team, 2024)\footnote{We, furthermore, used the R-packages \emph{bayesplot} (Version 1.11.1; Gabry, Simpson, Vehtari, Betancourt, \& Gelman, 2019), \emph{brms} (Version 2.21.0; Bürkner, 2017, 2018, 2021), \emph{citr} (Version 0.3.2; Aust, 2019), \emph{cmdstanr} (Version 0.8.1.9000; Gabry, Češnovar, Johnson, \& Bronder, 2024), \emph{dplyr} (Version 1.1.4; Wickham, François, Henry, Müller, \& Vaughan, 2023), \emph{forcats} (Version 1.0.0; Wickham, 2023a), \emph{ggplot2} (Version 3.5.1; Wickham, 2016), \emph{lme4} (Version 1.1.35.5; Bates, Mächler, Bolker, \& Walker, 2015), \emph{lubridate} (Version 1.9.3; Grolemund \& Wickham, 2011), \emph{Matrix} (Version 1.7.0; Bates, Maechler, \& Jagan, 2024), \emph{nlme} (Version 3.1.166; Pinheiro \& Bates, 2000), \emph{papaja} (Version 0.1.2.9000; Aust \& Barth, 2023), \emph{patchwork} (Version 1.2.0; Pedersen, 2024), \emph{purrr} (Version 1.0.2; Wickham \& Henry, 2023), \emph{RColorBrewer} (Version 1.1.3; Neuwirth, 2022), \emph{Rcpp} (Eddelbuettel \& Balamuta, 2018; Version 1.0.12; Eddelbuettel \& François, 2011), \emph{readr} (Version 2.1.5; Wickham, Hester, \& Bryan, 2024), \emph{RJ-2021-048} (Bengtsson, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021), \emph{standist} (Version 0.0.0.9000; Girard, 2024), \emph{stringr} (Version 1.5.1; Wickham, 2023b), \emph{tibble} (Version 3.2.1; Müller \& Wickham, 2023), \emph{tidybayes} (Version 3.0.6; Kay, 2023), \emph{tidyr} (Version 1.3.1; Wickham, Vaughan, \& Girlich, 2024), \emph{tidyverse} (Version 2.0.0; Wickham et al., 2019), and \emph{tinylabels} (Version 0.2.4; Barth, 2023).} for all reported analyses. Web links are printed in bold. The content of the tutorials is mainly based on Allison (2010), Singer and Willett (2003), McElreath (2018), Kurz (2023a), and Kurz (2023b).

@@

Figure 5 shows the effects of congruent and incongruent primes relative to neutral primes, for each time bin in trial number 1000 for the selected model.

{\centering \includegraphics[width=0.8\linewidth,height=0.67\textheight,]{../Tutorial_2_Bayesian/figures/M4effects_con_incon_trial1000}

\caption{50/80/95 percentile intervals of the draws from the posterior distributions representing the effect of congruent and incongruent primes relative to neutral primes in trial number 1000.}\label{fig:plot-prime-effects}
\end{figure}

Figure 6 shows the model-based hazard functions for each prime type for participant 6, in trial 500, 1000, and 1500.

{\centering \includegraphics[width=0.8\linewidth,height=0.67\textheight,]{../Tutorial_2_Bayesian/figures/M4effects_subject6}

\caption{Model-based hazard functions for participant 6 in trial 500, 1000, and 1500.}\label{fig:plot-hazard-subject6}
\end{figure}

@@

In this third tutorial we illustrate how to fit a multilevel hazard regression model in the frequentist framework, for the data set used in the first tutorial. For illustration purposes, we only fitted model M3 using the function glmer() from the package lme4.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{model\_M3\_f }\OtherTok{\textless{}{-}} \FunctionTok{glmer}\NormalTok{(event }\SpecialCharTok{\textasciitilde{}} \DecValTok{1} \SpecialCharTok{+}\NormalTok{ condition}\SpecialCharTok{*}\NormalTok{period\_9 }\SpecialCharTok{+} 
\NormalTok{                           condition}\SpecialCharTok{*}\FunctionTok{I}\NormalTok{(period\_9}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{) }\SpecialCharTok{+} 
\NormalTok{                           condition}\SpecialCharTok{*}\FunctionTok{I}\NormalTok{(period\_9}\SpecialCharTok{\^{}}\DecValTok{3}\NormalTok{) }\SpecialCharTok{+}
\NormalTok{                           trial\_c}\SpecialCharTok{*}\NormalTok{period\_9 }\SpecialCharTok{+} 
\NormalTok{                           trial\_c}\SpecialCharTok{*}\FunctionTok{I}\NormalTok{(period\_9}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{) }\SpecialCharTok{+} 
\NormalTok{                           trial\_c}\SpecialCharTok{*}\FunctionTok{I}\NormalTok{(period\_9}\SpecialCharTok{\^{}}\DecValTok{3}\NormalTok{) }\SpecialCharTok{+}
\NormalTok{                           (}\DecValTok{1} \SpecialCharTok{+}\NormalTok{ condition}\SpecialCharTok{*}\NormalTok{period\_9 }\SpecialCharTok{+}
\NormalTok{                           condition}\SpecialCharTok{*}\FunctionTok{I}\NormalTok{(period\_9}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{) }\SpecialCharTok{+}
\NormalTok{                           condition}\SpecialCharTok{*}\FunctionTok{I}\NormalTok{(period\_9}\SpecialCharTok{\^{}}\DecValTok{3}\NormalTok{) }\SpecialCharTok{+}
\NormalTok{                           trial\_c}\SpecialCharTok{*}\NormalTok{period\_9 }\SpecialCharTok{+} 
\NormalTok{                           trial\_c}\SpecialCharTok{*}\FunctionTok{I}\NormalTok{(period\_9}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{) }\SpecialCharTok{+} 
\NormalTok{                           trial\_c}\SpecialCharTok{*}\FunctionTok{I}\NormalTok{(period\_9}\SpecialCharTok{\^{}}\DecValTok{3}\NormalTok{) }\SpecialCharTok{|}\NormalTok{ pid),}

              \CommentTok{\# control parameters, data set, and complementary log{-}log link function}
              \AttributeTok{control =} \FunctionTok{glmerControl}\NormalTok{(}\AttributeTok{optimizer =} \FunctionTok{c}\NormalTok{(}\StringTok{"nlminbwrap"}\NormalTok{),}
                                     \AttributeTok{optCtrl =} \FunctionTok{list}\NormalTok{(}\AttributeTok{maxfun=}\DecValTok{10000000}\NormalTok{)), }
              \AttributeTok{data=}\NormalTok{M3\_data, }
              \AttributeTok{family=}\FunctionTok{binomial}\NormalTok{(}\AttributeTok{link=}\StringTok{"cloglog"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

In Figure 7 we compere the parameter estimates of model M3 from brms() with those of glmer().

{\centering \includegraphics[width=0.8\linewidth,height=0.67\textheight,]{../Tutorial_3_Frequentist/comparison}

\caption{Parameter estimates for model M3 from brms() and glmer().}\label{fig:plot-comparison}
\end{figure}

Figure 7 confirms that the parameter estimates from both Bayesian and frequentist models are pretty similar. However, the random effects structure of model M3 was already too complex for the frequentist model as it did not converge and resulted in a singular fit. This is of course one of the reasons why Bayesian modeling has become so popular in the last years.

@@

The hazard models in this tutorial assume that there is one event of interest. For RT data, this event consitutes a single transition between an ``idle'' state and a ``responded'' state. However, in certain situations, more than one event of interest might exist. For example, an individual might transition back and forth between a ``healthy'' state and a ``depression'' state, before being absorbed in a final ``death'' state. When you have data on the timing of these transitions, one can apply multi-state models which generalize survival analysis to transitions between three or more states (refs).
Also, the predictor variables in this tutorial are time-invariant, i.e., their value did not change over the course of a trial. Another extension is to include time-varying predictors, i.e., predictors whose value can change across the time bins within a trial.
